Machine Learning Models for the Segmentation of E-coaching Text

### 2018-03-07 paper-amia-2018/manuscript.tex
update introduction

\noindent{\bf Abstract}
\textit{Poor eating habits, particularly low fruit and vegetable intake, is a growing, serious public health concern among young adults. An effective intervention is required to improve eating habits. E-coaching is an intervention technique where eCoaches use email to deliver motivation-enhancing coaching to encourage healthy eating. To identify effective communication strategies of healthy eating, automatic annotation of email exchange is required where a critical step is email segmentation. In this study, we transformed segmentation task into a classification task and developed several state-of-the-art machine learning models including Support Vector Machine, Naive Bayes, K-Nearest Neighbor (KNN), Recurrent Neural Networks by utilizing contextual, topic and punctuation mark features. Results indicate that KNN is the best model and achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'', respectively. This study has a great implication to accelerate the pace of identifying effective communication strategies linked to healthy eating.}
%identify individual text segments, which can be annotated directly with a classification model, and

\section*{Introduction}
Unhealthy eating habits, particularly low fruit and vegetable intake, is a growing, serious public health concern, particularly among young adults age 21-30, referred to as Generation Y (GenY)\cite{blanck2008trends,centers2007fruit}. This generation has adopted a lifestyle that involves eating accessible, ``no mess'', quick, ``grab and go'' foods\cite{nebeling2007still,brug1999application}. They mainly eat ``out'' and infrequently shop and prepare food, limiting access to fruit and vegetables (FV)\cite{nelson2009improving,larson2006food}. Unfortunately, less than one-third of US adults\cite{blanck2008trends,ogden2006prevalence} and only 20\% of GenY\cite{blanck2008trends,american2006american,thompson2005dietary} eat the recommended 5 servings of fruit and vegetables daily. Those in inner-city urban and rural settings have among the poorest eating habits\cite{blanck2008trends,centers2007fruit,ogden2006prevalence,american2006american,thompson2005dietary}. GenY's poor dietary practices placing them at high risk for obesity and many chronic diseases, such as type 2 diabetes, as well as declines in predicted health status and life expectancy. Thus, there is a need to develop effective interventions to improve GenY's eating habits.

GenY is a tech-savvy generation requiring an intervention matched to their mobile lifestyle. Growing numbers use the internet to access health information with the largest increases in internet access among low-income Americans, making the internet well-suited for health promotion intervention\cite{strecher2007internet}. MENU GenY\cite{alexander2017motivations} (Making Effective Nutrition Choices for Generation Y) is a technology-based public health intervention to encourage increased fruit and vegetable intake among GenY. A critical component of MENU GenY is personalized e-coaching. E-coaches use email to deliver motivation-enhancing coaching to encourage healthy eating, grounded in the principles of Motivational Interviewing (MI), an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009ten,miller2009toward}. Patient ``change talk'', statements of intrinsic motivation about their desire, ability, reasons, need for and commitment to behavior change, is an established mediator of health behavior change\cite{apodaca2009mechanisms}. Identifying specific communication strategies linked to behavior change and integrating these strategies into communication-based interventions (e.g., brief, motivation-enhancing interventions delivered in a variety of settings or public health initiatives) can increase these interventions' potency.

A major drawback of this research is the qualitative methods traditionally used to analyze the communication process which is resource-intensive, requiring an iterative process of human (subjective) interpretation of the text. Rapidly developing computational technologies, specifically, machine learning models, offer a unique opportunity to accelerate this process. Our research group has recently applied machine learning-based models to MI sessions \cite{hasan2016study,kotov2015interpretable}. A simple communication code scheme was automated to characterize patient communication and achieved accuracy comparable to human coders\cite{hasan2016study}. However, this study utilized e-coaching data which have different format and context compared to motivational interview transcripts. MI session data was prepared by transcribing audio conversation which involves a counselor, an adolescent, and a caregiver. On the other hand, e-coaching data was directly collected from the email responses between an eCoach and young adult. By utilizing this e-coaching data and leveraging innovative machine learning models, the ultimate goal of this research study is to fully automate the communication coding process in eCoach-patient communication to increases in fruit and vegetable intake.

However, a significant barrier to fully automate e-coaching is the unsegmented email exchange. E-coaching data comprised of email responses which need to be segmented into groups of MI behaviors for developing an automatic classification of clinical interactions. However, automatic segmentation of e-coaching intervention sessions is a challenging task due to the 2 important reasons. First, the email is an unstructured text that contains informal email exchange in non-traditional formats. Second, a text segment not necessarily belongs to the entire sentence or collection of sentences. One sentence can be segmented into several MI behaviors and vice versa. Figure~\ref{fig:text-segment} illustrates the segmentation of an e-coaching email exchange where the first sentence segmented into 2 different MI behaviors. On the other hand, fourth and fifth segments contain only one and multiple sentences, respectively.

@@

Previous studies mainly focus on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2016sentence} in the medical domain. Apostolova et al.\cite{apostolova2009automatic} applied SVM by utilizing word-vector cosine similarity metric combined with several heuristics to classify clinical report into semantic sections such as demographics, history, exam procedure, finding, impression, etc. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for the section classification. In 2009, Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing technique, terminology-based rule, and naive Bayesian score for identifying sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM exploiting with linear kernel and recurrent convolutional neural networks with prosodic, part of speech features and word embeddings, were trained by Kreuzthaler et al.\cite{kreuzthaler2015detection} and Griffis et al.\cite{griffis2016quantitative}, respectively, for the detection of sentence boundary. However, segmentation of clinical conversation in the context of electronically delivered interventions, in particular, segmentation of MI or e-coaching text into groups of MI behaviors, is ignored while relying on manual hand-coded approach. Therefore, this study introduces a novel approach and the authors are not aware of any other work this approach has been considered for the segmentation of e-coaching text.

@@

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required a large set of data for both classes. In this study, we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

Punctuation mark plays an important role in segmentation boundary detection, and large numbers of errors were encountered by the false positive of boundary identification. Similarly, additional information is the common reason for the classified original segment into multiple segments. For example, [need help from April].

>>>>

\noindent{\bf Abstract}
\textit{Communication science to understand clinical process like Motivational Interviews (MI) is limited by traditional qualitative coding methods. Qualitative coding of communication data is very resource-intensive and time-consuming process, requires auto coding. This study utilized e-coaching data where email is used to deliver motivation-enhancing coaching to encourage healthy eating. A critical step toward automatic annotation of communication coding process is the segmentation of text data. In this study, we transformed segmentation task into a classification task and developed several state-of-the-art machine learning models including Support Vector Machine, Naive Bayes, K-Nearest Neighbor (KNN), Recurrent Neural Networks by utilizing contextual, topic and punctuation mark features. Results indicate that KNN is the best model and achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'' classes, respectively. This study has a great implication to save money and accelerate the pace of identifying effective communication strategies.}

\section*{Introduction}
Communication science to understand clinical process like Motivational Interviews (MI) is limited by traditional qualitative coding methods. Motivational Interviewing (MI) is an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009ten,miller2009toward}. Patient ``change talk'', statements of intrinsic motivation about their desire, ability, reasons, need for and commitment to behavior change, is an established mediator of health behavior change\cite{apodaca2009mechanisms}. Qualitative coding of communication data has been traditionally performed manually with pre-defined codes by trained coders, which is a tedious resource-intensive task, requiring an iterative process of human (subjective) interpretation of the text.

Rapidly developing computational technologies, specifically, machine learning models, offer a unique opportunity to accelerate this process. A few studies were done for the automatic annotation of clinical interactions in the setting of MI intervention. Lacson et al.\cite{lacson2005automatic} applied AdaBoost\cite{freund1999short} classifier for annotating the interactions in hemodialysis phone dialog into 3 categories. Our research group has recently applied several machine learning-based models to MI sessions \cite{hasan2016study,kotov2015interpretable}. A simple communication code scheme was automated to characterize patient communication and achieved accuracy comparable to human coders\cite{hasan2016study}.

Experimental data utilized in the above MI studies were prepared by transcribing audio conversation which involves a counselor, a patient, and some cases a caregiver\cite{hasan2016study,kotov2015interpretable}. However, this study utilized e-coaching data which have different structure and context compared to clinical data. E-coaches use email to deliver motivation-enhancing coaching to encourage healthy eating, grounded in the principles of motivational interviewing. E-coaching data is comprised of email responses which are unsegmented, unlike more traditional dyadic clinical communication where segmentation occurs naturally due to its conversation nature. Discourse analysis on e-coaching text can help to understand the communication process in e-coaching text by revealing the socio-psychological characteristics of a patient\cite{siegfried1995therapeutic}. By utilizing e-coaching data and leveraging innovative machine learning models, the ultimate goal of this research study is to fully automate the communication coding process.

A significant barrier to fully automated the communication coding process is the segmentation of the text data. During traditional qualitative coding, coders determine where to segment text, that is, where one code ends and another begins. Automatic segmentation of e-coaching intervention sessions is a challenging task due to the 2 important reasons. First, the email is an unstructured text that contains informal email exchange in non-traditional formats. Second, a text segment does not necessarily belong to the entire sentence or collection of sentences. One sentence can be segmented into multiple MI behaviors or several sentences may represent a single MI code. Figure~\ref{fig:text-segment} illustrates the segmentation of an e-coaching email exchange where the first sentence segmented into 2 different MI behaviors. On the other hand, fourth and fifth segments contain only one and multiple sentences, respectively.

@@

Previous studies mainly focus on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2016sentence} in the medical domain. Apostolova et al.\cite{apostolova2009automatic} applied SVM by utilizing word-vector cosine similarity metric combined with several heuristics to classify clinical report into semantic sections such as demographics, history, exam procedure, finding, impression, etc. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for the section classification. In 2009, Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing technique, terminology-based rule, and naive Bayesian score for identifying sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM exploiting with linear kernel and recurrent convolutional neural networks with prosodic, part of speech features and word embeddings, were trained by Kreuzthaler et al.\cite{kreuzthaler2015detection} and Griffis et al.\cite{griffis2016quantitative}, respectively, for the detection of sentence boundary.

Recently, an online clinical intervention called MENU GenY\cite{alexander2017motivations} (Making Effective Nutrition Choices for Generation Y) was tested. MENU GenY is a technology-based public health intervention to encourage increased fruit and vegetable intake among young adult age 21-30, utilized personalized e-coaching data. The main focus of MENU GenY is to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical conversation in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still ignored while relying on manual hand-coded approach. Therefore, this study introduces a novel approach and the authors are not aware of any other work this approach has been considered for the segmentation of e-coaching text.

@@

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required longer sequence with a large set of data for both classes. In this study, maximum length of the input sequence was 3 and we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

Punctuation mark plays an important role in segmentation boundary detection, and large numbers of errors were encountered by the false positive of boundary identification. For example, ``xxxxx xxxxxx'' was segmented after the second sentence where a punctuation mark was encountered. Similarly, additional information is the common reason for the classified original segment into multiple segments. For instance, ``xxxx xxxx xxxx'' was segmented after second sentence where third sentence supports the MI code confirmed by first two sentences.
